Continuous Deployment and Schema Evolution \\in SQL Databases

### 2015-02-23 paper.tex
Processed last of the feedback.

When service windows are short or even non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers. To address this problem we aim to find or implement a tool which allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older applications still operating on the older version of the schema.

@@

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database or schema and start recording transactions. When the copying has completed, apply the schema changes to the copied database or schema. Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state. Once this state has been achieved, it must either be maintained by continuing to record and playback transactions, or make the switch from the old to the new database or schema.}
  \item{\textbf{Approach 2}: When using \textit{Blue-Green deployments}, two instances of the database are available. Take the inactive database and apply the schema operations to it. In the mean time these two databases need to be kept synchronized. This can be done by using approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from the application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and construct SQL queries which can operate correctly on both schemas.}
\end{itemize}

@@

\subsubsection{TableMigrator~\cite{TableMigrator}}
This is a tool which works like Openark Kit and Percona Toolkit. Instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding records in both tables are synchronized with each other by passing over all records in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock on the original table --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps the names of the original and ghost table before releasing the write lock. Since this final pass can be done quickly, any application executing queries on that table is only very briefly blocked.

@@

\begin{itemize}
  \item None of the tools used in practice support foreign keys.
  \item The tools discussed primarily focus on MySQL databases. \textit{OpenArk Kit} and \textit{Percona Toolkit} are suits of tools which focus exclusively on MySQL database administration.
  \item These tools are limited in the sense that they typically only support making small changes to the existing database schema. This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.
\end{itemize}

@@

\begin{itemize}
  \item{A solution should not put restrictions on how to deploy applications, or which methods of deployment can be used: \textit{Blue-Green deployments}, \textit{Rolling upgrades}, \textit{Canary releases}, etc.}
  \item{Clients connected to the database must be able to access the contents of the database according to an active version of the database schema of their choosing. Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen version of the database schema, but contain the same data as is present in all other ``views''.}
  \item{Versioning and rolling back are important for release engineers. A solution should support versioning the changes being made to the database schema, and be able to rollback to a previous version if the need arises.}
  \item{Foreign keys are an integral part of SQL databases and should be supported.}
\end{itemize}

@@

\begin{figure}%
    \centering
    \subfloat[Original database schema]{{\includegraphics[width=2.6in]{pre-migration-state.png} }}%
    \qquad
    \subfloat[Database schema in mixed-state]{{\includegraphics[width=2.6in]{in-migration-state.png} }}%
    \qquad
    \subfloat[New database schema]{{\includegraphics[width=2.6in]{post-migration-state.png} }}%
    \caption{Different stages of the database schema when modifying the ``books'' table.}%
    \label{fig:migration-stages}%
\end{figure}

>>>>

When service windows are short or non-existent, these two problems make releasing new versions of applications frequently problematic for release engineers. To address this problem we aim to find or implement a tool which allows its users to not only alter their database schema in a non-blocking fashion, but also do this without affecting older applications still operating on the older version of the schema.

@@

\begin{itemize}
  \item{\textbf{Approach 1}: Copy the database or schema and start recording transactions. When the copying has completed, apply the schema changes to the copied database or schema. Finally replay all the recorded database transactions on the copied database or schema to reach an equivalent state. Once this state has been achieved, it must either be maintained by continuing to record and playback transactions, or make the switch from the old to the new database or schema.}
  \item{\textbf{Approach 2}: When using \textit{Blue-Green deployments}, two instances of the database are available. Take the inactive database and apply the schema operations to it. In the mean time these two databases need to be kept synchronized. This can be done by using Approach 1, or somehow ensure that the table under change remains in a read-only state until the active and inactive databases have switched roles.}
  \item{\textbf{Approach 3}: The last is an approach where schema operations are performed separately from the application deployments using operations which are non-blocking. As a consequence the application must be able to deal with either the old or the new database schema being in use, and construct SQL queries which can operate correctly on both schemas.}
\end{itemize}

@@

\subsubsection{TableMigrator~\cite{TableMigrator}}
This tool works like Openark Kit and Percona Toolkit. Instead of using triggers to synchronize the two tables with each other,  TableMigrator requires tables to have an additional column which stores the date and time that record was last modified. Using this information corresponding records in both tables are synchronized with each other by passing over all records in multiple passes. Once a pass can be done relatively quickly, TableMigrator acquires an exclusive write lock on the original table --- blocking the entire table --- and proceeds to do a final pass. Once this has been done, it atomically swaps the names of the original and ghost table before releasing the write lock. Since this final pass can be done quickly, any application executing queries on that table is only very briefly blocked.

@@

\begin{itemize}
  \item None of the tools used in practice support foreign keys. These tools will not work when a foreign key is present on a table under change. This forces software engineers to choose between not using any of these tools, or deliberately avoid defining foreign key constraints in their SQL databases in order to be able to use these tools.
  \item The tools discussed primarily focus on MySQL databases. \textit{OpenArk Kit} and \textit{Percona Toolkit} are suits of tools which focus exclusively on MySQL database administration.
  \item These tools are limited in the sense that they typically only support making small changes to the existing database schema. This forces software and release engineers to often release intermediate versions of the application, as the schema is altered in several steps.
\end{itemize}

@@

\begin{itemize}
  \item{A solution should not put restrictions on how to deploy applications, or which methods of deployment can be used: \textit{Blue-Green deployments}, \textit{Rolling upgrades}, \textit{Canary releases}, etc.}
  \item{Clients connected to the database must be able to access the contents of the database according to an active version of the database schema of their choosing. Their ``view'' should only contain tables, columns, foreign keys, etc which are present in their chosen version of the database schema.}
  \item{Versioning and rolling back are important for release engineers. A solution should support versioning the changes being made to the database schema, and be able to rollback to a previous version if the need arises.}
  \item{Foreign keys are an integral part of SQL databases and should be supported.}
\end{itemize}

@@

\begin{figure}%
    \centering
    \subfloat[Original database schema]{{\includegraphics[width=2.75in]{pre-migration-state.png} }}%
    \qquad
    \subfloat[Database schema in mixed-state]{{\includegraphics[width=2.75in]{in-migration-state.png} }}%
    \qquad
    \subfloat[New database schema]{{\includegraphics[width=2.75in]{post-migration-state.png} }}%
    \caption{Different stages of the database schema when modifying the ``books'' table.}%
    \label{fig:migration-stages}%
\end{figure}
